---
title: "Seabirds-Migration-Proteomics"
author: "Alix de Jersey and Jack Rivers-Auty"
date: "`r Sys.Date()`"
output:
    html_document:
      toc: true
      toc_title: "Table of Contents"
      number_sections: true
---

```{r CSS install, include=T, results='asis', comment=NA, echo=F}
x <- tryCatch(readLines("https://raw.githubusercontent.com/JackAuty/RmarkdownStyle/main/styles.css"), error=function(e) ""); if (length(x)) cat("<style>\n", paste(x, collapse="\n"), "\n</style>")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,       # Show code
  warning = FALSE,   # Suppress warnings
  message = FALSE    # Suppress messages
)

options(scipen =999 )
```

# Load Required Packages

```{r load-packages, warning=FALSE, message=FALSE, results='hide' }
# Load packages, install if required
packages <- c(
  "ggplot2", "reshape2", "enrichR", "stringr", "rgl", "htmltools", "pkgconfig",
  "PupillometryR", "viridis", "pheatmap", "MASS", "sjPlot", "emmeans", "remotes",
  "ggfortify", "RColorBrewer", "readr", "caret", "dplyr", "ggvenn",
  "ggVennDiagram", "pROC", "kableExtra", "tidyr", "DT")

invisible(lapply(packages, function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
  library(pkg, character.only = TRUE)
}))
```
# Citations
```{r citations}

# Use the existing package list
pkg <- packages 

# Create a character vector for clean citations
citation_list <- character()  

# Loop through each package and clean citation
for (i in pkg) {
  # Get the plain text citation
  raw <- paste(capture.output(print(citation(i), style = "text", bibtex = FALSE)), collapse = "\n")
  
  # Clean: remove _, *, " and trailing . or ,.
  clean <- gsub("_", "", raw, fixed = TRUE)
  clean <- gsub("\\*", "", clean)
  clean <- gsub('\\"', "", clean)            # Handles escaped quotes
  clean <- gsub('"', "", clean)              # Handles regular quotes
  clean <- gsub("\\\\texttt\\{(.*?)\\}", "\\1", clean)
  clean <- gsub(", \\.$", "", clean)
  clean <- gsub("\\s*\\.$", "", clean)
  clean <- gsub("[\u201C\u201D]", "", clean)
  clean <- trimws(clean)

  # Append to citation list
  citation_list <- c(citation_list, clean)
}

# Display as a table
datatable(
  data.frame(References = citation_list)
)
```

Requires two files to run: 

  1. Data file (Available through ProteomeXchange PXD064350)
  2. Sample information

```{r import data}
# Load in proteomic data
chicks_adults_df<- read_csv("github_data.csv")
# Load in sample information
chicks_adults_sampleinfo <- read_csv("github_chicks_adults_sampleinfo.csv")
```

```{r load age }
# Extract the 'Age' column from the sample information dataframe and assign it to a new variable 'age'
age <- chicks_adults_sampleinfo$Age
```

```{r clean proteins}
# Remove rows with missing or blank gene symbols
chicks_adults_df <- chicks_adults_df %>%
  filter(!is.na(gene_symbol), gene_symbol != "")
```

# Figure 1a. PCA Plot. 
```{r PCA}
col<-rep("#22a884", length(age)) # Assign a colour to all data points
col[age=="A"]<-"#355f8d" # Assign a different colour to one of the two groups. 

sum(!complete.cases(chicks_adults_df[, 5:47])) # Determine the number of proteins detected in all individuals.

# Transpose a complete (no NAs) matrix of the data columns for the PCA
PCA_matrix <- t(as.matrix(chicks_adults_df[complete.cases(chicks_adults_df[, 5:47]), 5:47])) 

PCA<-prcomp((PCA_matrix), scale=T) # Perform PCA on PCA_matrix with scaling = T (TRUE)
summary_PCA<-round(data.frame(summary(PCA)$importance),3) # Summarise each component
summary_PCA$PC1[2] # Extract the percentage variance explained by the second principal component (PC2)
#pdf("PCA Proteomics Biomarker FFSH Adults vs Plastic load.pdf", height = 7, width =8) # Set up a PDF file
par(fig=c(0, 1, 0, 1), oma=c(5, 2, 2, 2), mar=c(6, 4, 2, 2)) # Plot parameters and margins
plot(PCA$x[,1:2], col=col,pch=16, cex=1.5,xlab="",ylab="") # Create a scatterplot of the first two principal components
title(main="PCA Proteomics Biomarker FFSH Adults vs Low plastic chicks", # Add a title to the plot
      xlab=paste0("PCA1 ",(100*summary_PCA$PC1[2]),"%"), # Add axis titles to the plot
      ylab=paste0("PCA2 ",(100*summary_PCA$PC2[2]),"%"))
      legend("topleft", # Add a legend to the top left corner
              bty = "n", 
              legend = c("Fledglings", "Adults"), 
              pch = 16, 
              col = c(levels(as.factor(col))), 
              cex = 1, horiz=F, 
      title="Age")
```

# Statistics 

Calculate number of data points missing
```{r calculate missing data}
# Add empty columns to store the number of missing values for chicks and adults
chicks_adults_df$NumberMissingChicks <- NA
chicks_adults_df$NumberMissingAdults <- NA

for(i in 1:nrow(chicks_adults_df)){
  
  # Extract the data for the current row
  dat <- chicks_adults_df[i, 5:47]
  
  # Subset the data based on the group (Age: chicks, adults)
  chick <- dat[age == "F"]
  adult <- dat[age== "A"]
  
  # Calculate the number of missing values for each group
  chicks_adults_df$NumberMissingChicks[i] <- sum(is.na(chick))
  chicks_adults_df$NumberMissingAdults[i] <- sum(is.na(adult))
}

```

T test using the lm function with age as the explanatory variable with boxcox applied
```{r perform statistical analysis}
# Add empty columns to store the results of the statistical tests
chicks_adults_df$ttest <- NA
chicks_adults_df$LogFold <- NA

# Loop over each row of the dataframe
for(i in 1:nrow(chicks_adults_df)){
  print(i)  # Print the current row index for tracking progress
  dat <- as.numeric(chicks_adults_df[i, 5:47])  # Extract the data for each row
  
  # Assign data points to the respective groups
  chick <- dat[age == "F"]
  adult <- dat[age == "A"]
  
  # Calculate the number of non-missing data points for each group
  non_missing_chick <- sum(!is.na(chick))
  non_missing_adult <- sum(!is.na(adult))
  
  # Check if there are enough data points to perform the analysis
  if(non_missing_chick >= 14 & non_missing_adult >= 13) {
    
    # Fit the general linear model with age as the explanatory variable
    lm1 <- lm(dat ~ age)
    
    # Box-Cox transformation to normalize the data
    boxcox_result <- MASS::boxcox(lm1, lambda = seq(-5, 5, 0.003), plotit = FALSE)
    
    # Select the lambda that maximizes the Box-Cox log-likelihood
    Selected.lambda <- boxcox_result$x[which.max(boxcox_result$y)]
    
    # Apply the Box-Cox transformation to the data
    transDat <- dat^Selected.lambda
    
    # Perform the t-test on the transformed data
    t_test_result <- t.test(transDat ~ age)
    
    # Store the p value from the t-test in the new column for the current row
    chicks_adults_df$ttest[i] <- t_test_result$p.value
    
    # Calculate the log fold change between chick and adult age groups
    chicks_adults_df$LogFold[i] <- log(mean(adult, na.rm = TRUE) / mean(chick, na.rm = TRUE), base = 2)
    
  } else {
    # If not enough data points, assign NA to the t-test result and LogFold
    chicks_adults_df$ttest[i] <- NA
    chicks_adults_df$LogFold[i] <- NA
    print(i)
    print("Too few data points to perform stats")
  }
}

# Adjust for multiple comparisons
chicks_adults_df$p_adjust <- p.adjust(chicks_adults_df$ttest, method = "fdr")

# Calculate the proportion of missing data
chicks_adults_df$ProportionMissingChicks <- chicks_adults_df$NumberMissingChicks/19
chicks_adults_df$ProportationMissingAdults <- chicks_adults_df$NumberMissingAdults/24
```

```{r view statistical results}
# Sum the number of signficant protein pre and post adjustment
sum(chicks_adults_df$ttest<0.05, na.rm = TRUE)
sum(chicks_adults_df$p_adjust<0.01, na.rm = TRUE)
```

# Figure 1b. Volcano Plot. 
```{r volcano plot}
# Create a new column 'Sig' indicating whether the adjusted p-value is less than 0.05. 
chicks_adults_df$Sig <- ifelse(is.na(chicks_adults_df$p_adjust), FALSE, chicks_adults_df$p_adjust < 0.05)

# Create columns 'Up' and 'Down' indicating increased or decreased abundance of proteins based on Logfold and Sig
chicks_adults_df$Up <- chicks_adults_df$Sig & chicks_adults_df$LogFold > 0
chicks_adults_df$Down <- chicks_adults_df$Sig & chicks_adults_df$LogFold < 0

# Default grey for all points
chicks_adults_df$colour <- "#999999"

# Colour significant upregulated proteins (positive LogFold)
chicks_adults_df$colour[chicks_adults_df$Up] <- "#355f8d"

# Colour significant downregulated proteins (negative LogFold)
chicks_adults_df$colour[chicks_adults_df$Down] <- "#22a884"

# Create a new column 'order' with a default value of 0.95 for all proteins.
chicks_adults_df$order <- 0.95

# Adjust point size: larger for significant, smaller for not
chicks_adults_df$order <- ifelse(chicks_adults_df$Sig, 1.4, 0.95)

# Order the dataframe for plotting
df_graph_ordered <- chicks_adults_df[order(chicks_adults_df$order, decreasing = FALSE), ]

# Set plot limits
xmin <- -4
xmax <- 4
ymin <- 0.001
ymax <- 20

# Generate a volcano plot 
plot.new()
par(mfrow = c(1,1))

plot(df_graph_ordered$LogFold, 
     -log10(df_graph_ordered$p_adjust), 
     col = adjustcolor(df_graph_ordered$colour, alpha.f = 0.8), 
     pch = 16,
     cex = df_graph_ordered$order,
     xlab = "",
     ylab = "",
     xlim = c(xmin, xmax),
     ylim = c(ymin, ymax),
     axes = FALSE)

# Add axes
axis(1)
axis(2)

# Add labels
title(xlab = "Fold change (log2)", ylab = "Adjusted p-value (log10)", mgp = c(2,2,0))

# Add horizontal line for significance threshold (p = 0.05)
abline(h = -log10(0.05), lty = 3)

```

# Enrichment analysis
```{r subset control group}
# Isolate the datacols
ffsh2024_datacolumn <- 5:47

# Create a new data frame containing the adults which represent our control for the enrichment analysis.
enrichment_control_data<-chicks_adults_df[,-ffsh2024_datacolumn[age=="A"]]
```

```{r control enrichment}
# Extract the 'gene_symbol' column from the control enrichment results and store it as 'control_genes'
control_genes<-enrichment_control_data$gene_symbol

# Create a subset of the databases of interest
databases_of_interest<-c(
  "GO_Molecular_Function_2025", 
  "GO_Cellular_Component_2025", 
  "GO_Biological_Process_2025",
  "KEGG_2021_Human", 
  "WikiPathway_2021_Human")

# Perform enrichment using enrichr on the control genes  
enrichment_control<-enrichr(as.character(control_genes), databases = databases_of_interest)
```

```{r significant enrichment}
# Subset the gene symbols of the significant genes (where the p_adjust is less than 0.05) from the trimmed dataframe.
genes_of_significance <- chicks_adults_df$gene_symbol[chicks_adults_df$p_adjust < 0.05 
                                                              & !is.na(chicks_adults_df$p_adjust)]
# Create a subset of the databases of interest
databases_of_interest<-c("GO_Molecular_Function_2025",
                         "GO_Cellular_Component_2025", 
                         "GO_Biological_Process_2025",
                         "KEGG_2021_Human", 
                         "WikiPathway_2021_Human", 
                         "TRRUST_Transcription_Factors_2019")

#Selecting our genes and uploading them to 6 different enrichment databases  
enrichment_Sig<-enrichr(as.character(genes_of_significance), databases = databases_of_interest)
```

1.GO_Molecular_function_2025 database 
```{r G0_MF enrichment}
# Building the significant and control protein data frames for the GO_Molecular_Function database
go_mol_function_Sig <- enrichment_Sig$GO_Molecular_Function_2025[, c("Term", "Overlap", "Genes")]
go_mol_function_Control <- enrichment_control$GO_Molecular_Function_2025[, c("Term", "Overlap", "Genes")]

# Building empty columns for the loop
go_mol_function_Sig$PermutationTest <- 999
go_mol_function_Sig$ProteinNumber <- 999
go_mol_function_Sig$ProteinNumberControl <- 999
go_mol_function_Sig$TotalControlProtein <- nrow(enrichment_control_data)
go_mol_function_Sig$TotalSignificant <- sum(chicks_adults_df$p_adjust < 0.05 & !is.na(chicks_adults_df$p_adjust))

# Loop through each row of the significant data frame 
for (i in 1:nrow(go_mol_function_Sig)) {
  
  # Extract and process data from the "Overlap" column
  go_mol_function_Sig$ProteinNumber[i] <- as.numeric(sub("\\/.*", "", go_mol_function_Sig$Overlap[i])) 
  
  # Check if the term is present in the control data frame 
  if (go_mol_function_Sig$Term[i] %in% go_mol_function_Control$Term) {
    
    Control_data <- go_mol_function_Control[go_mol_function_Control$Term == go_mol_function_Sig$Term[i], ] 
    
    # Extract relevant information from the control data frame
    go_mol_function_Sig$ProteinNumberControl[i] <- as.numeric(sub("\\/.*", "", Control_data$Overlap)) 
    
  } else {
    
    # If the term is not present in the control data, set ProteinNumberControl to 0 
    go_mol_function_Sig$ProteinNumberControl[i] <- 0 
    
  }
}

# Trimming out the very low count pathways. There's no point saying a pathway is enriched with 2 genes
go_mol_function_Sig <- go_mol_function_Sig[go_mol_function_Sig$ProteinNumber > 2, ]

# Generate p-values by performing a permutation test
for (i in 1:nrow(go_mol_function_Sig)) {
  
  # Create a control vector for permutation test
  control <- c(
    rep(1, go_mol_function_Sig$ProteinNumberControl[i]), 
    rep(0, go_mol_function_Sig$TotalControlProtein[i] - go_mol_function_Sig$ProteinNumberControl[i])
  ) 
  
  # Perform a permutation test with 10,000 repeats
  repeats <- 10000 
  n <- rep(9999, repeats)
  
  for (j in 1:repeats) {
    n[j] <- sum(sample(control, go_mol_function_Sig$TotalSignificant[i]))  
  }
  
  # Calculate the p.value based on the permutation test
  significant <- go_mol_function_Sig$ProteinNumber[i] 
  go_mol_function_Sig$PermutationTest[i] <- sum(n >= significant) / repeats
  
  # Create a data frame for pathway and non-pathway counts 
  dat <- data.frame(
    "Control" = c(go_mol_function_Sig$ProteinNumberControl[i], 
                  (go_mol_function_Sig$TotalControlProtein[i] -       go_mol_function_Sig$ProteinNumberControl[i])),
    "Significant" = c(significant, 
                      (go_mol_function_Sig$TotalSignificant[i] - significant)),
    row.names = c("Pathway", "Non-pathway"),
    stringsAsFactors = FALSE
  ) 
  
  colnames(dat) <- c("Control", "Significant")
}

# Filter for significant pathways based on PermutationTest < 0.05 and view them
adult_go_mol_function_Sig <- go_mol_function_Sig[go_mol_function_Sig$PermutationTest < 0.05, ]
datatable(adult_go_mol_function_Sig)
```

2. GO_Cellular_Component_2025 database
```{r GO_CC enrichment}

GO_Cellular_Component_Sig<-enrichment_Sig$GO_Cellular_Component_2025[,c("Term","Overlap","Genes")]
GO_Cellular_Component_Control<-enrichment_control$GO_Cellular_Component_2025[,c("Term","Overlap","Genes")]

GO_Cellular_Component_Sig$PermutationTest<-999
GO_Cellular_Component_Sig$ProteinNumber<-999
GO_Cellular_Component_Sig$ProteinNumberControl<-999
GO_Cellular_Component_Sig$TotalControlProtein<-nrow(enrichment_control_data)
GO_Cellular_Component_Sig$TotalSignificant<-sum(chicks_adults_df$p_adjust < 0.05 &
                                                  !is.na(chicks_adults_df$p_adjust))

for(i in 1:nrow(GO_Cellular_Component_Sig)) {

  GO_Cellular_Component_Sig$ProteinNumber[i]<-as.numeric(sub("\\/.*", "", GO_Cellular_Component_Sig$Overlap[i]))
  
  if(GO_Cellular_Component_Sig$Term[i] %in% GO_Cellular_Component_Control$Term){
  Control_data<-GO_Cellular_Component_Control[GO_Cellular_Component_Control$Term==GO_Cellular_Component_Sig$Term[i],]
  
  GO_Cellular_Component_Sig$ProteinNumberControl[i]<-as.numeric(sub("\\/.*", "", Control_data$Overlap))
  
  } else {
    GO_Cellular_Component_Sig$ProteinNumberControl[i]<-0
  }
}

GO_Cellular_Component_Sig<- GO_Cellular_Component_Sig[GO_Cellular_Component_Sig$ProteinNumber>2,]

for(i in 1:nrow(GO_Cellular_Component_Sig)){

  control<-c(rep(1,GO_Cellular_Component_Sig$ProteinNumberControl[i]), rep(0,GO_Cellular_Component_Sig$TotalControlProtein[i] - 
                                                                             GO_Cellular_Component_Sig$ProteinNumberControl[i]))

  repeats<-10000

  n<-rep(9999, repeats)
  sum(control)

for (j in 1:repeats){
  
  n[j]<-sum(sample(control, GO_Cellular_Component_Sig$TotalSignificant[i]))
  
}

  significant<-GO_Cellular_Component_Sig$ProteinNumber[i]

  GO_Cellular_Component_Sig$PermutationTest[i]<- sum(n>=(significant))/repeats

  dat <- data.frame(
    "Control" = c(GO_Cellular_Component_Sig$ProteinNumberControl[i], (GO_Cellular_Component_Sig$TotalControlProtein[i] -
                                                                      GO_Cellular_Component_Sig$ProteinNumberControl[i])),
    "Significant" = c(significant, (GO_Cellular_Component_Sig$TotalSignificant[i]-significant)),
    row.names = c("Pathway", "Non-pathway"), stringsAsFactors = FALSE
)
  colnames(dat) <- c("Control", "Significant")

  dat
}


adult_GO_Cellular_Component_Sig <- GO_Cellular_Component_Sig[GO_Cellular_Component_Sig$PermutationTest< 0.05, ]
datatable(adult_GO_Cellular_Component_Sig)
```

3. GO_Biological_Process_2025 database
```{r GO_BP enrichment}
GO_Bioloigcal_Process_Sig<-enrichment_Sig$GO_Biological_Process_2025[,c("Term","Overlap","Genes")]
GO_Bioloigcal_Process_Control<-enrichment_control$GO_Biological_Process_2025[,c("Term","Overlap","Genes")]

GO_Bioloigcal_Process_Sig$PermutationTest<-999
GO_Bioloigcal_Process_Sig$ProteinNumber<-999
GO_Bioloigcal_Process_Sig$ProteinNumberControl<-999
GO_Bioloigcal_Process_Sig$TotalControlProtein<-nrow(enrichment_control_data)
GO_Bioloigcal_Process_Sig$TotalSignificant<-sum(chicks_adults_df$p_adjust < 0.05 &
                                                  !is.na(chicks_adults_df$p_adjust))

for(i in 1:nrow(GO_Bioloigcal_Process_Sig)) {

  GO_Bioloigcal_Process_Sig$ProteinNumber[i]<-as.numeric(sub("\\/.*", "", GO_Bioloigcal_Process_Sig$Overlap[i]))
  
  if(GO_Bioloigcal_Process_Sig$Term[i] %in% GO_Bioloigcal_Process_Control$Term){
  Control_data<-GO_Bioloigcal_Process_Control[GO_Bioloigcal_Process_Control$Term==GO_Bioloigcal_Process_Sig$Term[i],]
  
  GO_Bioloigcal_Process_Sig$ProteinNumberControl[i]<-as.numeric(sub("\\/.*", "", Control_data$Overlap))
  
  } else {
    GO_Bioloigcal_Process_Sig$ProteinNumberControl[i]<-0
  }
}

GO_Bioloigcal_Process_Sig<- GO_Bioloigcal_Process_Sig[GO_Bioloigcal_Process_Sig$ProteinNumber>2,]

for(i in 1:nrow(GO_Bioloigcal_Process_Sig)){

  control<-c(rep(1,GO_Bioloigcal_Process_Sig$ProteinNumberControl[i]), rep(0,GO_Bioloigcal_Process_Sig$TotalControlProtein[i] - 
                                                                             GO_Bioloigcal_Process_Sig$ProteinNumberControl[i]))

  repeats<-10000

  n<-rep(9999, repeats)
  sum(control)

for (j in 1:repeats){
  
  n[j]<-sum(sample(control, GO_Bioloigcal_Process_Sig$TotalSignificant[i]))
  
}

  significant<-GO_Bioloigcal_Process_Sig$ProteinNumber[i]

  GO_Bioloigcal_Process_Sig$PermutationTest[i]<- sum(n>=(significant))/repeats

  dat <- data.frame(
    "Control" = c(GO_Bioloigcal_Process_Sig$ProteinNumberControl[i], (GO_Bioloigcal_Process_Sig$TotalControlProtein[i] -
                                                                      GO_Bioloigcal_Process_Sig$ProteinNumberControl[i])),
    "Significant" = c(significant, (GO_Bioloigcal_Process_Sig$TotalSignificant[i]-significant)),
    row.names = c("Pathway", "Non-pathway"), stringsAsFactors = FALSE
)
  colnames(dat) <- c("Control", "Significant")

  dat
}

#View(GO_Bioloigcal_Process_Sig)

adult_GO_Bioloigcal_Process_Sig <- GO_Bioloigcal_Process_Sig[GO_Bioloigcal_Process_Sig$PermutationTest < 0.05, ]
datatable(adult_GO_Bioloigcal_Process_Sig)
```

4. KEGG_2021_Human
```{r GO_KEGG enrichment}
KEGG_Sig<-enrichment_Sig$KEGG_2021_Human[,c("Term","Overlap","Genes")]
KEGG_Control<-enrichment_control$KEGG_2021_Human[,c("Term","Overlap","Genes")]

KEGG_Sig$PermutationTest<-999
KEGG_Sig$ProteinNumber<-999
KEGG_Sig$ProteinNumberControl<-999
KEGG_Sig$TotalControlProtein<-nrow(enrichment_control_data)
KEGG_Sig$TotalSignificant<-sum(chicks_adults_df$p_adjust < 0.05 & !is.na(chicks_adults_df$p_adjust))

for(i in 1:nrow(KEGG_Sig)) {

  KEGG_Sig$ProteinNumber[i]<-as.numeric(sub("\\/.*", "", KEGG_Sig$Overlap[i]))
  
  if(KEGG_Sig$Term[i] %in% KEGG_Control$Term){
  Control_data<-KEGG_Control[KEGG_Control$Term==KEGG_Sig$Term[i],]
  
  KEGG_Sig$ProteinNumberControl[i]<-as.numeric(sub("\\/.*", "", Control_data$Overlap))
  
  } else {
    KEGG_Sig$ProteinNumberControl[i]<-0
  }
}

KEGG_Sig<- KEGG_Sig[KEGG_Sig$ProteinNumber>2,]

for(i in 1:nrow(KEGG_Sig)){

  control<-c(rep(1,KEGG_Sig$ProteinNumberControl[i]), rep(0,KEGG_Sig$TotalControlProtein[i] - KEGG_Sig$ProteinNumberControl[i]))

  repeats<-10000

  n<-rep(9999, repeats)
  sum(control)

for (j in 1:repeats){
  
  n[j]<-sum(sample(control, KEGG_Sig$TotalSignificant[i]))
}

  significant<-KEGG_Sig$ProteinNumber[i]

  KEGG_Sig$PermutationTest[i]<- sum(n>=(significant))/repeats

  dat <- data.frame(
    "Control" = c(KEGG_Sig$ProteinNumberControl[i], (KEGG_Sig$TotalControlProtein[i] - KEGG_Sig$ProteinNumberControl[i])),
    "Significant" = c(significant, (KEGG_Sig$TotalSignificant[i]-significant)),
    row.names = c("Pathway", "Non-pathway"), stringsAsFactors = FALSE
)
  colnames(dat) <- c("Control", "Significant")

  dat
}

#View(KEGG_Sig)

adult_KEGG_Sig <- KEGG_Sig[KEGG_Sig$PermutationTest < 0.05, ]
datatable(adult_KEGG_Sig)
```

5. WikiPathway_2021_Human database 
```{r WikiPathway enrichment}
WikiPathway_Sig<-enrichment_Sig$WikiPathway_2021_Human[,c("Term","Overlap","Genes")]
WikiPathway_Control<-enrichment_control$WikiPathway_2021_Human[,c("Term","Overlap","Genes")]

WikiPathway_Sig$PermutationTest<-999
WikiPathway_Sig$ProteinNumber<-999
WikiPathway_Sig$ProteinNumberControl<-999
WikiPathway_Sig$TotalControlProtein<-nrow(enrichment_control_data)
WikiPathway_Sig$TotalSignificant<-sum(chicks_adults_df$p_adjust < 0.05 & !is.na(chicks_adults_df$p_adjust))

for(i in 1:nrow(WikiPathway_Sig)) {

  WikiPathway_Sig$ProteinNumber[i]<-as.numeric(sub("\\/.*", "", WikiPathway_Sig$Overlap[i]))
  
  if(WikiPathway_Sig$Term[i] %in% WikiPathway_Control$Term){
  Control_data<-WikiPathway_Control[WikiPathway_Control$Term==WikiPathway_Sig$Term[i],]
  
  WikiPathway_Sig$ProteinNumberControl[i]<-as.numeric(sub("\\/.*", "", Control_data$Overlap))
  
  } else {
    WikiPathway_Sig$ProteinNumberControl[i]<-0
  }
}

WikiPathway_Sig<- WikiPathway_Sig[WikiPathway_Sig$ProteinNumber>2,]

for(i in 1:nrow(WikiPathway_Sig)){

  control<-c(rep(1,WikiPathway_Sig$ProteinNumberControl[i]), rep(0,WikiPathway_Sig$TotalControlProtein[i] - 
                                                                   WikiPathway_Sig$ProteinNumberControl[i]))

  repeats<-10000

  n<-rep(9999, repeats)
  sum(control)

for (j in 1:repeats){
  
  n[j]<-sum(sample(control, WikiPathway_Sig$TotalSignificant[i]))
}

  significant<-WikiPathway_Sig$ProteinNumber[i]

  WikiPathway_Sig$PermutationTest[i]<- sum(n>=(significant))/repeats

  dat <- data.frame(
    "Control" = c(WikiPathway_Sig$ProteinNumberControl[i], (WikiPathway_Sig$TotalControlProtein[i] - 
                                                            WikiPathway_Sig$ProteinNumberControl[i])),
    "Significant" = c(significant, (WikiPathway_Sig$TotalSignificant[i]-significant)),
    row.names = c("Pathway", "Non-pathway"), stringsAsFactors = FALSE
)
  colnames(dat) <- c("Control", "Significant")

  dat
}

#View(WikiPathway_Sig)

adults_WikiPathway_Sig <- WikiPathway_Sig[WikiPathway_Sig$PermutationTest <0.05, ]
datatable(adults_WikiPathway_Sig)
```

# Figure 2. Coagulation heatmap
```{r coagulation heatmap}
# Define gene set related to coagulation pathways
coagulation <- c("C7", "C8B", "CD55", "F10", "F11", "F2", "FGA", "FGB", "FGG", "VWF") 

# Subset to rows where gene_symbol is in 'coagulation', and select:
# - protein abundance data (columns 5:47)
# - gene_symbol (column 55)
# - adjusted p-value (column 49)
pre_heatmap_df <- chicks_adults_df[chicks_adults_df$gene_symbol %in% coagulation, c(5:47, 54, 49)]

# For each gene, keep only the row with the lowest p_adjust (most significant entry)
pre_heatmap_df <- pre_heatmap_df %>%
  group_by(gene_symbol) %>%
  slice_min(order_by = p_adjust, n = 1, with_ties = FALSE) %>%
  ungroup()

# Apply a square root transformation to the protein abundance values
pre_heatmap_df[, 1:length(age)] <- sqrt(pre_heatmap_df[, 1:length(age)])

# Generate a heatmap:
# - data: transformed abundance values
# - row labels: gene symbols
# - column labels: ordered by age (F for chicks, A for adults)
# - no clustering of rows or columns
# - rows scaled (Z-score)
# - RdBu color palette reversed for better visual contrast
pheatmap(pre_heatmap_df[, 1:length(age)],
         labels_row = pre_heatmap_df$gene_symbol,
         scale = "row",
         labels_col = age[c(which(age == "F"), which(age == "A"))],
         fontsize_row = 12,
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         legend = TRUE,
         breaks = seq(-2, 2, length.out = 11),
         cellwidth = 20,      # width of each sample column
         cellheight = 20,     # height of each gene row
         color = rev(brewer.pal(11, "RdBu")))
```

Function to graph individual proteins of interest for data exploration
# Figure 3. Raincloud plot 
```{r graph protein function}
graph_protein <- function(gene_symbol) {
  # Subset the data columns for the specified gene
  protein_Count <- chicks_adults_df[chicks_adults_df$gene_symbol  == gene_symbol , 5:47]
  
  # Sum protein counts across samples, keeping NAs
  protein_Count <- colSums(protein_Count, na.rm = FALSE)
  
  # Create a new dataframe for plotting
  graph_data <- data.frame(age, as.numeric(protein_Count))
  colnames(graph_data) <- c("Age", "Protein_count")
  
  # Count missing values for annotation (A = adults, F = fledglings)
  missing <- c(
    sum(is.na(graph_data$Protein_count[graph_data$Age == "A"])),
    sum(is.na(graph_data$Protein_count[graph_data$Age == "F"]))
  )
  
  # Define colorblind-friendly green (young) and blue (adult)
  colours <- c("F" = "#22a884", "A" = "#355f8d")
  
  # Generate the raincloud plot
  rain_cloud_plot <- ggplot(graph_data, aes(Age, Protein_count, fill = Age, col = Age)) + 
    geom_flat_violin(position = position_nudge(x = 0.1, y = 0), alpha = 1, width = 0.5) +
    geom_point(position = position_jitterdodge(dodge.width = 0.15, jitter.width = 0.15), size = 2, alpha = 1) +
    geom_boxplot(color = "black", width = 0.1, outlier.shape = NA, alpha = 1, 
                 position = position_nudge(x = -0.2, y = 0)) +
    coord_flip() +
    labs(x = "", y = "Relative levels", 
         title = paste("", gene_symbol)) +  # Add protein name to title
    scale_y_continuous(trans = "sqrt") +
    theme_classic() +
    theme(legend.position = "none", text = element_text(size = 14)) +
    scale_fill_manual(values = colours) +
    scale_colour_manual(values = colours) +
    scale_x_discrete(labels = c("A" = "Adult", "F" = "Fledgling")) +
    annotate("text", label = missing, x = levels(as.factor(graph_data$Age)), 
             y = rep(min(graph_data$Protein_count, na.rm = TRUE)/4, 2)) +
    annotate("text", label = "Not detected", x = 0.5, 
             y = min(graph_data$Protein_count, na.rm = TRUE)/3) +
    geom_hline(yintercept = min(graph_data$Protein_count, na.rm = TRUE)*3/4,
               linetype = 'dashed', col = "#666666", linewidth = 1)
  
  print(rain_cloud_plot)
}
```

```{r graphing proteins of interest, warning=FALSE, message=FALSE, echo=FALSE}
# Insert protein symbol of choice
graph_protein("")
```

# Figure 4. Metabolism heatmap
```{r metabolism heatmap}
# Define a vector of gene symbols related to metabolic pathways
metabolism <- c("ENO1", "PGM1", "TPI1", "APOD", "FABP4", "ALDOB", "MINPP1", "PKM", "ME1") 

# Subset the dataframe to only rows with genes in 'metabolism' and select relevant columns:
# protein abundance data (cols 5:47), gene symbol (col 55), and p_adjust (col 49)
pre_heatmap_df <- chicks_adults_df[chicks_adults_df$gene_symbol %in% metabolism, c(5:47, 54, 49)]

# For genes with multiple entries, keep only the row with the lowest p_adjust (most significant)
pre_heatmap_df <- pre_heatmap_df %>%
  group_by(gene_symbol) %>%
  slice_min(order_by = p_adjust, n = 1, with_ties = FALSE) %>%
  ungroup()

# Convert gene_symbol to a factor to retain the desired order specified in 'metabolism'
pre_heatmap_df$gene_symbol <- factor(pre_heatmap_df$gene_symbol, levels = metabolism)

# Reorder the dataframe rows based on the factor levels of gene_symbol
pre_heatmap_df <- pre_heatmap_df[order(pre_heatmap_df$gene_symbol), ]

# Apply a square root transformation to the abundance data (columns 1 to length of 'age')
pre_heatmap_df[, 1:length(age)] <- sqrt(pre_heatmap_df[, 1:length(age)])

# Generate a heatmap:
# - input: transformed abundance data
# - labels for rows: gene symbols (ordered)
# - labels for columns: sample age (F for chicks, A for adults)
# - scaling applied by row
# - no clustering
# - customized cell size, breaks, and color scale
pheatmap(pre_heatmap_df[, 1:length(age)],
         labels_row = pre_heatmap_df$gene_symbol,
         scale = "row",
         labels_col = age[c(which(age == "F"), which(age == "A"))],
         fontsize_row = 12,
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         legend = TRUE,
         cellwidth = 20,
         cellheight = 20,
         breaks = seq(-2, 2, length.out = 11),
         color = rev(brewer.pal(11, "RdBu")))
```

Backwards enrichment analysis
```{r backwards enrichment analysis}
#install.packages("BiocManager")
#BiocManager::install("org.Hs.eg.db")

library("org.Hs.eg.db")
go_term <- "GO:0062023" # Insert pathway of interest
columns(org.Hs.eg.db)
genes <- select(org.Hs.eg.db, 
                keys = go_term,
                keytype = "GO",
                columns = c("SYMBOL", "GENENAME", "ONTOLOGY")
                )
col_ecm <- genes
```

# Figure S1. Collagen-Containing Extracellular Matrix (GO:0062023) heatmap 
```{r collagen heatmap}
# Extract gene symbols from a dataframe of extracellular matrix-related genes
colecm <- col_ecm$SYMBOL

# Subset chicks_adults_df to ECM-related genes, selecting abundance data (cols 5:47) and adjusted p-values (col 49)
pre_heatmap_df <- chicks_adults_df[chicks_adults_df$gene_symbol %in% colecm, c(5:47, 49)]

# Apply square root transformation to the protein abundance values
pre_heatmap_df[, 1:length(age)] <- sqrt(pre_heatmap_df[, 1:length(age)])

# Remove rows with any missing values (complete cases only)
pre_heatmap_df_trim <- pre_heatmap_df[complete.cases(pre_heatmap_df), ]

# Generate heatmap of ECM-related genes (trimmed and clustered by rows)
pheatmap(pre_heatmap_df_trim[, 1:length(age)],             # Input matrix: transformed abundance values
         labels_row = pre_heatmap_df_trim$gene_symbol,     # Label rows by gene symbol
         scale = "row",                                    # Scale values per row (Z-score transformation)
         labels_col = age[c(which(age == "F"), which(age == "A"))],  # Label columns by age group: chicks ("F") and adults ("A")
         fontsize_row = 12,                                # Row label font size
         cluster_cols = FALSE,                             # Do not cluster columns (samples)
         cluster_rows = TRUE,                              # Cluster rows (genes)
         legend = TRUE,                                    # Show color legend
         breaks = seq(-2, 2, length.out = 11),             # Define color breakpoints
         color = rev(brewer.pal(11, "RdBu")))              # Use reversed Red-Blue color palette: blue = lower, red = higher expression
```

# Figure S2. Adults maturation
```{r adult maturation}
# Subset the sample info to adult individuals only (rows 20 to 43)
adults_sampleinfo <- chicks_adults_sampleinfo[20:43, ]


# Subset the protein expression data: include p_adjust (col 49) and protein abundances (cols 24 to 47)
ffsh_adults <- chicks_adults_df[, c(49, 24:47)]

# Extract the Maturity classifications as a vector
maturity <- adults_sampleinfo$Adult_Age

# Assign colors: default is dark blue for all, then light blue for Immature
col <- rep("#0072b2", length(maturity))           # Default color for all
col[maturity == "Immature"] <- "#56b4e9"          # Change color for Immature group

# Create PCA input matrix:
# - Select only columns 2:25 (protein abundance)
# - Use only rows with no missing values (complete cases)
# - Transpose so proteins are rows, samples are columns
PCA_matrix <- t(as.matrix(ffsh_adults[complete.cases(ffsh_adults), 2:25]))

# Perform Principal Component Analysis with scaling
PCA <- prcomp(PCA_matrix, scale = TRUE)

# Summarize PCA importance (variance explained per component), rounded to 3 decimal places
summary_PCA <- round(data.frame(summary(PCA)$importance), 3)

# Check the percentage of variance explained by PC1 and PC2
summary_PCA$PC1[2]  # PC1 variance
summary_PCA$PC2[2]  # PC2 variance

# Optional: Uncomment to save as PDF
# pdf("PCA Proteomics Biomarker FFSH Adults vs Plastic load.pdf", height = 7, width = 8)

# Set up plot margins and figure layout
par(fig = c(0, 1, 0, 1), oma = c(5, 2, 2, 2), mar = c(6, 4, 2, 2))

# Create scatterplot of PCA scores (first two principal components)
plot(PCA$x[, 1:2],
     col = col,
     pch = 16,
     cex = 1.5,
     xlab = "",
     ylab = "")

# Add title and axis labels with % variance explained
title(main = "PCA FFSH adults maturity",
      xlab = paste0("PCA1 ", (100 * summary_PCA$PC1[2]), "%"),
      ylab = paste0("PCA2 ", (100 * summary_PCA$PC2[2]), "%"))

# Add a legend indicating maturity groups
legend("topleft",
       bty = "n",  # no box
       legend = c("Immature", "Mature"),
       pch = 16,
       col = c("#56b4e9", "#0072b2"),  # match to assigned colors
       cex = 1,
       horiz = FALSE,
       title = "Age")

```
